---
title: "Final Report"
author: "Kayla Howa"
date: "`r Sys.Date()`"
output:
  github_document:
  toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Read in libraries
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(plyr)

## Read in data set
apples <- read.csv(file="appleHarvest.csv")
apples$city <- NULL

## Change categorical variables to numeric variables
apples$harvest <- revalue(apples$harvest, c("Minimal"="1", "Normal"="2", "Above average"="3"))
apples$harvest <- as.numeric(apples$harvest)

## Change Yes and No to single letters
apples$pruned <- revalue(apples$pruned, c("Yes"="Y", "No"="N"))
apples$pruned <- as.factor(apples$pruned)

apples$elevation <- as.factor(apples$elevation)

apples$harvest
summary(apples$harvest)

na.omit(apples)
```

# ABSTRACT


# STUDY QUESTION & HYPOTHESIS

## Question
Does pruning trees lead to an increase in fruit yield in apple trees?
  
## Hypothesis
I hypothesize that pruning trees leads to an increase in fruit yield in apple trees.

## Prediction
I predict that by pruning fruit bearing trees, such as apples, the fruit yield will increase for the following harvest season.

# METHODS

### Method 1


### Method 2


### Method 3


### Analysis

## First Analysis
```{r first-analysis, warning=FALSE}
##Chi-Square Test
chisq.test(apples$pruned, apples$harvest)

##ANOVA

```

## Second Analysis
```{r second-analysis, eval=TRUE, fig.align='left', fig.height=7, fig.width=4.75, warning=FALSE, include=TRUE}
## Creating a linear model to try and determine whether pruning impacts apple harvest abundance
apple_lm <- lm(harvest ~ pruned + elevation, data = apples)
appleResids <- apple_lm$residuals
qqnorm(appleResids)

appleANOVA <- aov(apple_lm)
summary(appleANOVA)

TukeyHSD(appleANOVA, "elevation")

```

## Third Analysis
```{r, third-analysis, warning=T}
counts <- table(apples$harvest, apples$pruned)
summary(counts)
print(counts)

barplot(counts, main = "Distribution of Harvest Quantity by Tree Pruning", 
        cex.sub = 0.75, sub = "Total Number of Observations = 242",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))

```

# DISCUSSION

### Interpretation - First Analysis

### Interpretation - Second Analysis


### Limitations



# CONCLUSION


# REFERENCES

