---
title: "Final Report"
author: "Kayla Howa"
date: "`r Sys.Date()`"
output:
  github_document:
  toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Read in libraries
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(plyr)
library(rstatix)

## Read in data set
apples <- read.csv(file="appleHarvest.csv")
apples$city <- NULL
na.omit(apples)

## Change categorical variables to numeric variables
apples$harvest <- revalue(apples$harvest, c("Minimal"="1", "Normal"="2", "Above average"="3"))
apples$harvest <- as.numeric(apples$harvest)

## Change Yes and No to single letters
apples$pruned <- revalue(apples$pruned, c("Yes"="Y", "No"="N"))
apples$pruned <- as.factor(apples$pruned)

apples$elevation <- as.factor(apples$elevation)
apples$harvest
summary(apples$harvest)


```

# ABSTRACT

# BACKGROUND
Apples held the top spot for total fruit available for consumption in 2021 (Kantor and Blazejczyk, 2023) and, according to USApple’s analysis, total U.S. apple production for the 2025/26 crop year (CY) is forecast at 11.7 billion pounds / 278.5 million bushels (USApple, 2025).\ 

With apples being such a hot commodity, it is important to figure out what factors may be influencing harvest sizes. We have been provided a dataset 

# STUDY QUESTION & HYPOTHESIS

## Question

Does pruning trees lead to an increase in fruit yield in apple trees?

## Hypothesis

I hypothesize that pruning trees leads to an increase in fruit yield in
apple trees.

## Prediction

I predict that by pruning fruit bearing trees, such as apples, the fruit
yield will increase for the following harvest season.

# METHODS


### Method 1
We have fit a simple linear model to the data, with y being the response
variable (harvest amount) and x~1~ and x~2~ being our covariate factors
(whether the tree was pruned and tree elevation). Because the pruned covariate is categorical, dummy variables were automatically created for it. One pruned category is left out in order to prevent collinearity among the data. 

The final model statement for y is:

$$y = \beta_{0} + \beta_{1}x_{1} + \beta_{2}x_{2} + \beta_{3}x_{3} + \epsilon$$
$$\epsilon \sim N(0,\sigma^2)$$

where $\beta$ describes the effects of each predictor on y and
$\epsilon$ are normally distributed residual error terms with a variance
of $\sigma^2$.



### Method 2
We performed a Chi-Square test of independence between our two
variables pruned and harvest

Where pruned is represented as either Yes or No, and

Where harvest is represented as 1, 2, or 3: 1 = Minimal 2 = Average 3 =
Above Average


### Method 3

### Analysis

## First Analysis
```{r first-analysis, eval=TRUE, fig.align='left', fig.height=7, fig.width=4.75, warning=FALSE, include=TRUE}
## Create scatterplot matrix to determine relationships
plot(~ pruned + harvest + elevation + year, data = apples)

##Create scatterplot to see relationships
plot(apples$harvest, apples$pruned)
plot(apples$harvest, apples$elevation)
plot(apples$harvest, apples$year)
     
## Creating a linear model to try and determine whether pruning impacts apple harvest abundance
apple_lm <- lm(harvest ~ pruned + elevation + year, data = apples)
summary(apple_lm)
appleResids <- apple_lm$residuals
qqnorm(appleResids)
shapiro.test(appleResids)


```

## Second Analysis
```{r second-analysis, warning=FALSE}
## ANOVA
options(max.print=1000000)
appleANOVA <- aov(apple_lm)
summary(appleANOVA)

TukeyHSD(appleANOVA, which = "elevation")


```

## Third Analysis
```{r, third-analysis, warning=T}
## Pruned Plots
countsPruned <- table(apples$harvest, apples$pruned)
summary(countsPruned)
print(countsPruned)

barplot(countsPruned, main = "Distribution of Harvest Quantity by Tree Pruning", 
        cex.sub = 0.75, sub = "Total Number of Observations = 242",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))


##Elevation Plots
countsElevation <- table(apples$harvest, apples$elevation)
summary(countsElevation)
print(countsElevation)

barplot(countsElevation, main = "Distribution of Harvest Quantity by Elevation", 
        cex.sub = 0.75, sub = "Total Number of Observations = 144",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))

##Year Plots
countsYear <- table(apples$harvest, apples$year)
summary(countsYear)
print(countsYear)

barplot(countsYear, main = "Distribution of Harvest Quantity by Elevation", 
        cex.sub = 0.75, sub = "Total Number of Observations = 144",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))

```

# DISCUSSION

### Interpretation - First Analysis

### Interpretation - Second Analysis

### Limitations

Issues with Alternate bearing in addition to a lack of year to year data. 

Missing data within rows.

Poorly labeled values
  -> what does 'normal' 'above average' and 'minimal' mean?
  -> what are they comparing it to?
  -> what does it mean?

# CONCLUSION

# REFERENCES
2025-26 apple production will reach nearly 279 million bushels. USApple. (2025, August 15). https://usapple.org/news-resources/2025-26-apple-crop-outlook 

Kantor, L., & Blazejczyk, A. (2023, May 5). Apples and oranges are the top U.S. fruit choices. Apples and oranges are the top U.S. fruit choices | Economic Research Service. https://www.ers.usda.gov/data-products/chart-gallery/chart-detail?chartId=58322 
