---
title: "Final Report"
author: "Kayla Howa"
date: "`r Sys.Date()`"
output:
  github_document:
  toc: true
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#Read in libraries
library(ggplot2)
library(ggpubr)
library(RColorBrewer)
library(plyr)
library(rstatix)
library(DescTools)


## Read in data set
apples <- read.csv(file="appleHarvest.csv")
apples$city <- NULL
na.omit(apples)

## Change categorical variables to numeric variables
apples$harvest <- revalue(apples$harvest, c("Minimal"="1", "Normal"="2", "Above average"="3"))
apples$harvest <- as.numeric(apples$harvest)

## Make elevation and year factors for ANOVA testing
apples$elevation <- as.factor(apples$elevation)
apples$year <- as.factor(apples$year)


```

# ABSTRACT



# BACKGROUND
Apples held the top spot for total fruit available for consumption in
2021 (Kantor and Blazejczyk, 2023) and, according to USApple’s analysis,
total U.S. apple production for the 2025/26 crop year (CY) is forecast
at 11.7 billion pounds / 278.5 million bushels (USApple, 2025). 

With apples being such a hot commodity, it is important to figure out
what factors may be influencing harvest sizes to ensure bountiful
harvests season after season. We have been provided a dataset from the
Wasatch Back Fruit Tree Project, a citizen led research project through
Utah State University Extension. Their goal is to assemble data on the
fruit tree varieties that can reliably produce on the Wasatch Front
(About the Wasatch, n.d.).

# STUDY QUESTION & HYPOTHESIS
## Questions
The questions we are interested in looking into are:
1.How does elevation impact fruit yield?
2.Do older trees prodcue more fruit than younger trees?

## Hypothesis
We hypothesize that high elevation leads to an decrease in fruit yield in apple trees. Additionally, we hypothesize that older trees will produce more fruit than younger trees.

## Prediction
We believe that elevation will lead to a decrease in fruit yield due less CO2 in the air for the tree to consume and that more mature trees will produce more fruit because they rely on new growth less. 

# METHODS & ANALYSIS
### Method 1
We have created multiple bar plots to aid in visualization of the spread of the data. The bar plots are also useful to try to initially predict whether any of the factors we are looking into affect harvest quantity. 

### First Analysis
```{r, first-analysis, warning=T}
##Elevation Plots
countsElevation <- table(apples$harvest, apples$elevation)
summary(countsElevation)
print(countsElevation)

barplot(countsElevation, main = "Distribution of Harvest Quantity by Elevation", 
        cex.sub = 0.75, sub = "Total Number of Observations = 144",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))

##Year Plots
countsYear <- table(apples$harvest, apples$year)
summary(countsYear)
print(countsYear)

barplot(countsYear, main = "Distribution of Harvest Quantity by Elevation", 
        cex.sub = 0.75, sub = "Total Number of Observations = 144",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))

```


### Method 2
We have fit a simple linear regression model to the data for each factor to determine if either one may be impacting fruit yield. For each of the two models, y is the response variable (harvest amount) and x~1~ is the covariate factor that we are analyzing (year or elevation).

The final model statement for each y is:

$$y = \beta_{0} + \beta_{1}x_{1} + \epsilon$$
$$\epsilon \sim N(0,\sigma^2)$$

where $\beta$ describes the effects of each predictor on y and $\epsilon$ are normally distributed residual error terms with a variance of $\sigma^2$.

### Second Analysis
```{r second-analysis, eval=TRUE, fig.align='left', fig.height=7, fig.width=4.75, warning=FALSE, include=TRUE}
## Creating a linear model for each factor to try and determine impact on fruit harvest
elevation_lm <- lm(harvest ~ elevation, data = apples)
summary(elevation_lm)

year_lm <- lm(harvest ~ year, data = apples)
summary(year_lm)
```

### Method 3
Because both the elevation linear model and the year linear model show potential 

### Third Analysis
```{r third-analysis, warning=FALSE, fig.align='center', fig.height=25, fig.width=7,}
yearANOVA <- aov(year_lm)
summary(yearANOVA)

year_Tukey <- TukeyHSD(yearANOVA)
print(year_Tukey)

elevationANOVA <- aov(elevation_lm)
summary(elevationANOVA)
```

# DISCUSSION

### Interpretation - First Analysis


### Interpretation - Second Analysis



### Interpretation - Third Analysis
Due to our ANOVA test for year being highly significant, but seeing no significance in the Tukey's HSD, we can conclude that one or more population means are different from each other, but we cannot draw a confident conclusion about which ones. There was no need to run a Tukey's HSD for elevation, as the ANOVA returned no signifance.




### Limitations
There are a few limitations that come with this data set. The first one is that apple trees may produce a heavy crop of fruit in one year and then very little to almost none the next. This is called “alternate” or “biennial bearing”. The typical pattern is 2 alternating years, but some trees may have an excessive crop one year and then very little for the next two years (University of Saskatchewan, 2018). The lack of knowledge about what part of this alternate bearing cycle the dataset trees are in limits us to not be able to make conclusive causation claims with any of our tested factors.

The second limitation is with the dataset itself. There are many observations that are missing data within rows. The amount of data is inconsistent. Additonanlly, the values in the data set are poorly labeled. For example, harvest amount in the dataset is labeled as "normal", "minimal", and "above average". We don't have a reference about what any of these are supposed to suggest or what the harvest amounts are being compared to.

# CONCLUSION

# REFERENCES

2025-26 apple production will reach nearly 279 million bushels. USApple.
(2025, August 15).
<https://usapple.org/news-resources/2025-26-apple-crop-outlook>

About the Wasatch Back Fruit Tree Project. Utah State University.
(n.d.). <https://extension.usu.edu/wasatch/WBFTP_about>

Kantor, L., & Blazejczyk, A. (2023, May 5). Apples and oranges are the
top U.S. fruit choices. Apples and oranges are the top U.S. fruit
choices \| Economic Research Service.
<https://www.ers.usda.gov/data-products/chart-gallery/chart-detail?chartId=58322>

University of Saskatchewan. (2018, February 2). Alternate or biennial bearing on fruit trees. College of Agriculture and bioresources. https://gardening.usask.ca/articles-and-lists/articles-disorders/disorder-alternate-or-biennial-bearing-on-fruit-trees.php  

Wikimedia Foundation. (2025, October 22). Dummy Variable (statistics). Wikipedia. https://en.wikipedia.org/wiki/Dummy_variable_(statistics) 

ChatGPT. OpenAI, version Jan 2025.



