---
title: "Final Report"
author: "Kayla Howa"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#Read in libraries
library(ggplot2)
library(RColorBrewer)
library(plyr)

## Read in data set
apples <- read.csv(file="appleHarvest.csv")
apples$elevation <- NULL
apples$city

## Change categorical variables to numeric variables
apples$harvest <- revalue(apples$harvest, c("Minimal"="1", "Normal"="2", "Above average"="3"))
apples$harvest <- as.numeric(apples$harvest)


## Change Yes and No to single letters
apples$pruned <- revalue(apples$pruned, c("Yes"="Y", "No"="N"))
apples$pruned <- as.factor(apples$pruned)



apples$harvest
summary(apples$harvest)

```

# ABSTRACT
NA = 0
minimal = 1
normal = 2
above average = 3
prune = y
no prune = n

# STUDY QUESTION & HYPOTHESIS

## Question
Does pruning trees lead to an increase in fruit yield in apple trees?


## Hypothesis
We hypothesize that pruning trees leads to an increase in fruit yield in apple trees.

## Prediction


## 

# METHODS

### Method 1


### Method 2


### Method 3



### Analysis



## First Analysis
```{r first-analysis, warning=FALSE}
## one-sample t-test 


```

## Second Analysis

```{r second-analysis, eval=TRUE, fig.align='left', fig.height=7, fig.width=4.75, warning=FALSE, include=TRUE}
## Creating a linear model to try and determine whether pruning impacts apple harvest abundance
apple_lm <- lm(harvest ~ pruned, data = apples)
resids <- apple_lm$residuals
qqnorm(resids)


```

## Third Analysis

```{r, third-analysis, warning=T}
counts <- table(apples$harvest, apples$pruned)

barplot(counts, main = "Distribution of Harvest Quantity by Tree Pruning", 
        cex.sub = 0.75, sub = "Total Number of Observations = 242",xlab = "Was the Tree Pruned?", 
        ylab = "Counts", ylim = c(0,100),col = c("#117733", "#44AA99", "#88CCEE"))

```

# DISCUSSION

### Interpretation - First Analysis

### Interpretation - Second Analysis


### Limitations



# CONCLUSION


# REFERENCES

